---
title: "Using Mass Spectrometry Data from MetaboLights"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Using Mass Spectrometry Data from MetaboLights}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{MsBackendMetaboLights}
    %\VignetteDepends{Spectra,BiocStyle}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("MsBackendMetaboLights")`<br />
**Authors**: `r packageDescription("MsBackendMetaboLights")[["Author"]] `<br />
**Last modified:** `r file.info("MsBackendMetaboLights.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r, echo = FALSE, message = FALSE}
library(Spectra)
library(BiocStyle)
```

# Introduction

The `r Biocpkg("Spectra")` package provides a central infrastructure for the
handling of Mass Spectrometry (MS) data. The package supports interchangeable
use of different *backends* to import MS data from a variety of sources and data
formats. The *MsBackendMetaboLights* package allows to retrieve MS data files
directly from the [MetaboLights](https://www.ebi.ac.uk/metabolights/)
repository. MetaboLights is one of the main public repositories for deposition
of metabolomics experiments including (raw) MS and/or NMR data files and the
related experimental and analytical results. The *MsBackendMetaboLights* package
downloads and locally caches MS data files for a MetaboLights data set and
enables further analyses of this data directly in R.


# Installation

The package can be installed directly from within R with the few commands below:

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RforMassSpectrometry/MsBackendMetaboLights")
```


# Importing MS Data from MetaboLights

[MetaboLights](https://www.ebi.ac.uk/metabolights/) is one of the main public
repositories for deposition of metabolomics experiments including (raw) mass
spectrometry (MS) and NMR data files and experimental/analysis results. The
experimental metadata and results are stored as plain text files in ISA-tab
format. Each MetaboLights experiment must provide a file describing the samples
analyzed and at least one *assay* file that links between the experimental
samples and the (raw and processed) data files with quantification of
metabolites/features in these samples.

In this vignette we explore and load MS data files from a small MetaboLights
experiment. MetaboLights provides information on a data set/experiment as a set
of plain text files in *ISA-tab* format. These can be accessed and read from the
data set's ftp folder. The set of files consist generally of a file with
information on the experiment/investigation (in a file with the file name
starting with *i_*) the samples of the data set (file name starting with *s_*),
the *assay* (measurements/analysis) of the experiment and a file with quantified
metabolite abundances (file name starting with *m_*). Note that a data set can
have more than one assay file.

Below we list all files from the MetaboLights data set with the ID *MTBLS39*.

```{r}
library(MsBackendMetaboLights)

#' List files of a MetaboLights data set
all_files <- mtbls_list_files("MTBLS39")
```

All these files are directly accessible in the ftp folder associated with the
MetaboLights data set. Below we use the `mtbls_ftp_path()` function to return
the ftp path for our test data set.

```{r}
mtbls_ftp_path("MTBLS39")
```

We could inspect the content of this folder also using a browser supporting the
ftp file transfer protocol and also download the files. We can however access
the files also directly from within R. Below we read the *assay* data file
directly using the `read.table()` function.

```{r}
#' Get the assay files of the data set
grep("^a_", all_files, value = TRUE)

#' Read the assay file
a <- read.table(paste0(mtbls_ftp_path("MTBLS39"),
                       grep("^a_", all_files, value = TRUE)),
                sep = "\t", header = TRUE, check.names = FALSE)
```

Each row in this assay table refers to one measurement (data file) of the data
set, with columns providing information on that measurement. The number and
content of columns can vary between data sets. Below we list the columns
available in the assay file of our test data set.

```{r}
colnames(a)
```

MS data files are generally provided in a column named `"Derived Spectral Data
File"` or `"Raw Spectral Data File"`, depending into which data field the data
set submitted entered the data. Below we list the content of these data columns.

```{r}
a[, c("Raw Spectral Data File", "Derived Spectral Data File")]
```

Note that providing MS data files is not absolutely mandatory, thus, for some
data sets no MS data files might be available. For this particular data set the
MS data files are provided in the `"Raw Spectral Data File"` column. These files
are in CDF format and can hence be loaded using the `MsBackendMetaboLights`
backend into R as a `Spectra` object (`MsBackendMetaboLights` directly extends
*Spectra*'s `MsBackendMzR` backend and therefore supports import of MS data
files in mzML, CDF or mzXML formats). By default, all MS data files of all
assays would be retrieved, but in our example below we restrict to few files to
reduce the amount of data that needs to be downloaded. We define for that a
pattern matching the file name of only some data files using the `filePattern`
parameter. Alternatively, for data sets with more than one assay, it would also
be possible to select MS data files from one particular assay only using the
`assayName` parameter. In our case we load all MS data files that end with
*63A.cdf*.

```{r}
library(Spectra)

#' Load MS data files of one data set
s <- Spectra("MTBLS39", filePattern = "63A.cdf",
             source = MsBackendMetaboLights())
s
```

This call now downloaded the files to the local cache and loaded these files as
a `Spectra` object. The downloading and caching of the data is handled by
Bioconductor's `r Biocpkg("BiocFileCache")`. Any subsequent loading of the same
data files will load the locally cached versions avoiding thus repetitive
download of the same data.

The message that is shown by the call above indicates that the MS data files
were not provided in the expected column (`"Derived Spectral Data File"`) but in
the column for raw data files.

The `Spectra` object with the MS data files of the MetaboLights data set enables
now any subsequent of the data that supports this type of data. On top of the
spectra variables and mass peak data values that are provided by the MS data
files also additional information related to the MetaboLights data set are
available as specific *spectra variables*. We list all available spectra
variables of the data set below.

```{r}
spectraVariables(s)
```

The MetaboLights-specific variables are `"mtbls_id"`, `"mtbls_assay_name"` and
`"derived_spectral_data_file"`.

```{r}
spectraData(s, c("mtbls_id", "mtbls_assay_name",
                 "derived_spectral_data_file"))
```

These variables can also be used to e.g. link the individual spectra back to the
original sample (e.g. through the *assay* and *sample* tables of the
MetaboLights data set.


# Session information

```{r}
sessionInfo()
```
